Neural networks for solving regression and classification problems

### 2022-11-16 doc/paper.tex
Add stuff

We developed a Feed Forward Neural Network (FFNN) and applied it to many common datasets used for benchmarking machine learning algorithms. Furthermore, we compared our results to those attained by linear and logistic regression. To optimize our network, we explored the use of different schedulers in our gradient descent method, and compared their performance against eachother. Overall, we found that our FFNN performed well in all classification tasks we tested, attaining test accuracies between $95$ and $100$ percent for all classification problems. However, we also found that simple logistic regression performed similarly, with a much lower computational cost. For regression on real terrain data, we found that our neural network slightly outperformed our previous polynomial fit, attaining an MSE of MONKEY, PERCENT percent lower than simple linear regression. In our gradient descent, we found that SCHEDULER performed best, being 1 BILLION percent better than the average of all other methods.

@@

% Neural Networks are all the rage these days, everyone wants to have a neural network in their home. Ive even seen people install neural networks in their dogs. \cite{ericjona}

% Neural networks are powerful tools, able to solve many different problems with

@@

The Franke Function is a simple weighted sum of polynomials, taking two inputs and returning a real value. It is commonly used to test machine learning algorithms for regression problems. For our linear regression problems, our training data will be based on a 9th degree polynomial approximation to the Franke Function, with 55 features. For the feed forward neural network, we will use 2 of the features (x, y) of the Franke Function.

\subsubsection{Elevation data over Norway}

This data is similar to the Franke Function in that it takes two inputs (x and y coordinates) and returns a real value, which is the elevation of a specific point in Norway.

@@

\begin{table}[h]
\caption{Characteristics of each dataset used in the report}
\begin{center}
\label{datasettable}
\begin{tabular}{| c | l | l | l |}
\hline
Dataset & Number of instances & Features & Type \\
\hline
Franke Function & synthetic & 55 & regression \\
Franke Function & synthetic & 2 & regression \\
Terrain (downscaled) & 204 800 & 2 & regression \\
Wisconsin Breast Cancer & 569 & 30 & binary classification \\
MNIST 8x8 & 1797 & 64 & multiclass classification (10 classes) \\
\hline
\end{tabular}
\end{center}
\end{table}

@@

In order to study the attributes of the different scheduler methods we performed numerical linear regression using gradient descent on the Franke Function with 55 features. We started off by comparing Newtons method for gradient descent to a constant learning rate using full batch gradient descent.

From observing FIG CONSTANT V HESSIAN we see that Newton's method using Hessian inversion greatly outperforms the constant scheduler by reaching an MSE of 0.02 within the first epoch. This is due to the Hessian inversion in the first epoch in Newton's method being equivalent to the analytical solution.  By comparison, the constant scheduler given 200 epochs reaches an MSE of 0.21. Given this result, it would seem indefensible to claim that Newtons method requires improvements, especially considering the time it took the Newton method 0.1512 seconds to compute 200 epochs, again beating the result from the constant scheduler which took 0.4991 seconds \footnote{this result should be taken with a grain of salt, given that the constant scheduler is run as a part of a larger architecture, whilst the Hessian calculation is done in a seperate script. Still, this provides an intuition that Newton's method does not take much time to compute this specific problem}. This result is due to the Hessian matrix only being a mere 55x55 matrix for this problem, such that the matrix inversion will not be computationally complex. However, if we were to increase the complexity of the problem, Newtons method would rapidly increase in computational complexity as it is a large matrix inversion with many floating point operations. For problems of such a small scale, when our features are based on a polynomial approximation Franke Function, it can be important to remember that we don't necessarily have to turn to these fancier solutions as the simple Newton method will do the job perfectly. Keeping this lesson in mind, we wish to further investigate the various scheduler methods in order to gain a better understanding of how well they perform, which will become useful for larger problems.

From FIGURE MOMENTUM v CONSTANT we can observe that adding momentum reduces the amount of epochs necessary to reach a lower MSE score. Already after EPOCHS epochs, the momentum scheduler has achieved MOMENTUM MSE AT EPOCHS, a notable improvement over CONSTANT MSE AT EPOCHS achieved by a constant learning rate.

By introducing mini-batch gradient descent of different batch sizes, we can observe from BATCH SIZE FIG that the scheduler methods perform better when optimized for different batch sizes. Both the constant and the momentum scheduler seem to vary equally over batch sizes, with the optimal batch size being equal to one. This means that our schedulers work the best for a fully stochastic gradient descent. The MSE with full gradient descent, as can be seen in BATCH SIZE FIG when batch size equals 320, provides an MSE of 
A full comparison of batch sizes for all the different schedulers can be viewed in APPENDIX in the appendix.

Furthermore, with the introduction of various adaptive scheduler methods such as RMS prop, adagrad, adagrad with momentum and adam we can observe from FIGURE ALL that we can reach even greater results in fewer epochs. We can also observe that the different schedulers all reach an MSE of 0.02 within a couple of epochs, but in order to get a better understanding of the computational cost we have also compared the time it took to fit the data to an MSE of 0.2. Looking at TIME TABLE FOR ALL SCHEDULERS, Adagrad took the least time, beating out our baseline constant scheduler by 0,222106 seconds. Based on FIG ALL, Adam seems to reach an MSE of 0.2 the fastest when strictly looking at the number of epochs required, however, whilst Adagrad requires more epochs the computations are computed faster in our implementation. Surprisingly, RMS prop performed more poorly than expected compared to other adaptive schedulers.

An interesting insight from FIGURE NUM is that all the schedulers prefer high learning rates and regularization. As this plot was generated for 100 epochs due to the computational complexity of gridsearching the six schedulers, the bias towards higher learning rates is certainly due to the high learning rates being able to reach a low MSE in the short training time. With the lambda = 0 column representing the ordinary least squares solutions, these are clearly consistantly outperformed by Ridge regression as the best MSE is achieved with regularization. For this specific plot, momentum does not seem to make out a noticeable difference for the best solutions, as the optimal solution equals their momentumless counterparts. That being said, in FIG ALL it is clear that the added momentum aids in a quicker convergence when compared to the constant scheduler.

Another observation from FIG NUM  is that the schedulers final error for100 epochs are more or less the same. As the largest change in MSE occurs in the first 50 epochs as seen in ALL FIGS, perhaps the other schedulers are given enough time to catch up as the best performing schedulers, such as Adagrad, starts slowing down the rate at which the MSE is decreased. However, as can be seen in HEATMAP FIG, Adagrad offers a 12\% \footnote{$\frac{constant MSE}{adagrad MSE} * 100 = \frac{0.018}{0.015} * 100 = 12\%$} decrease in MSE compared to the constant scheduler. It is clear from the results thus far that Adagrad is the best scheduler for approximating the Franke function when using linear regression with gradient descent. We believe that the reason Adagrad wins out is that these result are based on training for 100 epochs. The downfall of Adagrad in the literature seems to be that the aggressive decaying of the learning rate hinders a quick convergence deeper into training SOURCE https://towardsdatascience.com/learning-parameters-part-5-65a2f3583f7d. We are thus cautionary to claim that Adagrad will do the best job over a much larger amount of epochs when compared to adaptive schedulers that claim to improve this aspect.

To test this, we compared Adagrad with Adam over 5000 epochs. Both schedulers performed equally, both resulting in an MSE of 0.017 with Adagrad taking the edge with an MSE 0.0002 lower. Scikit MSE 0.005.

Furthermore, when trained for 5000 epochs, Adam with optimal parameters keeps improving the MSE. Scikit learn's Adam scheduler does perform better, offering a greater decrease the MSE. We compared Adagrad to scikit learn's Adam MLPRegressor as this seemed to be the closest option for comparing our results. Even though we have cross validated both our own Adagrad implementation and Scikit's with 5 folds, these differences in results might be due to the difference in solver, a more favorable weight initiation in Scikits model or additional optimizations present in the scikit learn implementation. Nonetheless, neither models are able to beat the analytical solution provided by linear regression methods using polynomial approximations, with our best result of an MSE of 0.012 on the test data and Scikits final MSE of 0.004 (granted this is on the train data, as scikit learn does not provide the loss curve for the test data).  both falling short of the analytical MSE of 0,003. In theory, the feed forward neural network should be able to better approximate the Franke function given that it is not limited to polynomial solutions. This result might be due to 5000 epochs not being enough training time, as the Franke function is a quite complex function with added noise.

@@

For the organic terrain data used in our previous project, we found that our neural net performed slightly better, gaining an MSE of NUMBER, which was about PERCENTAGE better than our best polynomial fit. However, looking at our prediction visually, and also comparing the MSE to the one gained on the Franke Function, we see that we are still lacking. For terrain data like this, which contain a lot of spatial information, we believe that Convolutional Neural Networks could be a better fit.

@@

To test our neural networks ability to solve classification problems, we first applied it to the Wisconsin Breast Cancer dataset. Using our gridsearch method, we found that the best scheduler was Adam, with an eta of $0.01$ and no regularization. For our architecture, we chose to use only one hidden layer with 60 nodes, based on our testing. With this, we were able to achieve a crossvalidated test accuracy of $98.6 \%$, which is very good. Looking at the confusion matrix (Fig.~\ref{cancerconf}), we see that we are achieving good predictions for both true positives and true negatives.

\begin{figure}[h]
\centerline{\includegraphics[width=5in]{figure/cancerconf.png}}
\caption{Confusion matrix for the cancer dataset using a neural net with a single hidden layer of 60 nodes}
\label{verification}
\end{figure}

Confident in our networks's ability to to binary classification, we turned to the MNIST dataset. Using our gridsearch, we again found that the best scheduler was Adam. Our architecture in this case was a single hidden layer with 100 nodes. Being a multiclass problem with ten different classes, one might expect our neural network to struggle more with this dataset. However, this was not the case, as we achieved a crossvalidated test accuracy of $99.07$. From these two datasets, we can see that a correctly optimized neural network is adept at solving classification problems. On the other hand, these good results, and the easew with which our network attained them, could imply that the problems we have dealt with are not as complicated as they may seem.

@@

In this paper we have explored the use of a Feed Forward Neural Network for both regression and classification. We have studied how different schedulers,

@@

\begin{table*}
\caption{This table shows the command to execute to reproduce every figure in the report. (More info about the scripts and their parameters can be found in the README)}
\begin{center}
\label{allparamstable}
\begin{tabular}{c | l l l}
Figure & Shell command (leading \texttt{python3} omitted) \\
\hline
\ref{frankenonoise} & \texttt{task\_b.py --noise 0 -n 25}\\
\hline
\end{tabular}
\end{center}
\end{table*}

>>>>

We developed a Feed Forward Neural Network (FFNN) and applied it to both regression and classification problems. Furthermore, we compared our results to those attained by linear and logistic regression. To optimize our network, we explored the use of different schedulers in our gradient descent method, and compared their performance against eachother. Amongst our schedulers, we found that Adam and Adagrad performed the best. Overall, we found that our FFNN performed well in both classification tasks we tested, attaining a test accuracy of $98.76$ on the Wisconsin Breast Cancer Dataset and a test accuracy of $99.07$ on the MNIST 8x8 dataset. However, we also found that simple logistic regression performed similarly, with a much lower computational cost. For regression applied to the Franke Function, we found that our neural net performed worse than a simple polynomial fit using linear regression, attaining a MSE of $0.007$, more than twice as high as the one achieved using linear regression.

@@

@@

The Franke Function is a simple weighted sum of polynomials, taking two inputs and returning a real value. It is commonly used to test machine learning algorithms for regression problems.

@@

\begin{table}[h]
\caption{Characteristics of each dataset used in the report}
\begin{center}
\label{datasettable}
\begin{tabular}{| c | l | l | l |}
\hline
Dataset & Number of instances & Features & Type \\
\hline
Franke Function & synthetic & 2 & regression \\
Wisconsin Breast Cancer & 569 & 30 & binary classification \\
MNIST 8x8 & 1797 & 64 & multiclass classification (10 classes) \\
\hline
\end{tabular}
\end{center}
\end{table}

@@

In order to study the attributes of the different scheduler methods we performed numerical linear regression using gradient descent on the Franke Function. IF HESSIAN: Using Newtons method, the computational cost is much greater than simply replacing the Hessian matrix inversion with a constant learning rate as can be seen by TIME COMPARISON. Furthermore, we can observe from FIGURE COMPARING HESSIAN TO CONSTANT that the constant scheduler does a good job at reducing the MSE, given that the final MSE score is CONSTANT SCORE compared to the Newton method's HESSIAN SCORE.

From FIGURE MOMENTUM v CONSTANT we can observe that adding momentum reduces the amount of epochs necessary to reach a lower MSE score. Already after EPOCHS epochs, the momentum scheduler has achieved MOMENTUM MSE AT EPOCHS, a notable improvement over CONSTANT MSE AT EPOCHS achieved by a constant learning rate. Furthermore, with the introduction of various adaptive scheduler methods such as RMS prop, adagrad, adagrad with momentum and adam we can observe from FIGURE ALL that we can reach even greater results in fewer epochs. We can also observe that the different schedulers that they all reach REASONABLE MSE within a couple of epochs, but in order to get a better understanding of the computational cost we have also compared the time it took to fit the data. Looking at TIME TABLE FOR ALL SCHEDULERS, BEST took the least time, beating out our baseline constant scheduler by TIME DIFF. Surprisingly, RMS prop performed more poorly than expected.

Before continuing, it is interesting to note that all the schedulers work best with a combination of similar eta and lambda values when gridsearched for achieving the minimal MSE. It is also interesting that they in the long term reach more or less the same error as can be observed in HEATMAP FIG. This could be because given more epochs, the other schedulers are given more time to catch up as the best performing schedulers, such as Adam, only decrease their MSE slightly. However, as can be seen in HEATMAP FIG, Adam offers a $\frac{0.018}{0.016} \dot 100 = 11.11\%$ decrease in MSE compared to the constant schedulers.

However, when trained for 500 epochs, Adam with optimal parameters does not yield any improvement in MSE. Scikits Adam scheduler does perform better, further decreasing the MSE slightly. Even though we have cross validated both our own Adam implementation and Scikit's with 5 folds, these differences in results might be due to a more favorable weight initiation in Scikits model. Nonetheless, neither models are able to beat the analytical solution provided by linear regression methods using polynomial approximations, with our best result of ADAM FINAL MSE and Scikits final MSE of SCIKIT FINAL MSE both falling short of the analytical MSE of 0.003. In theory, the feed forward neural network should be able to better approximate the Franke function given that it is not limited to polynomial solutions. This result might be due to 500 epochs not being enough training time.

@@

Having gained an understanding of how the different schedulers function, and having created the necessary infrastructure to study our models performance based on a number of different factors, we use these methods to optimize our neural network for linear regression. More specifically, we attempt to fit the Franke function. First, we perform a preliminary test on MSE over model complexity. For comparisons sake, we use the same scheduler for each architecture, and fix the regularization parameter at 0. By fixing the regularization parameter at 0, we are of course favouring less complex models, but this a good way to get a general sense of when the model starts to overfit. Using the sigmoid activation function, we can study the MSE as a function of the total hidden node count (Fig.~\ref{msebymodelfranke}). Here we see, somewhat surprisingly, that our train and test MSE are not deviating particularly much from eachother, even with very complex networks. There are many reasons that this may be the case. As we have stated before, once we start using numerical methods, computational effort becomes a large factor in any test one does, and many decisions have to made regarding how much time can be spent on each test. Due to these computational constraints, we have not been able to grid search for the best parameters for each model complexity, nor have we been able to train for more than 200 epochs for each model. These shortcomings will greatly affect how our models perform, as it is unlikely that each model uses the same learning rate and scheduler, or that more complex models use the same amount of epochs to train their weights. Based on this, we can assume that the higher complexity models have not been trained enough, and that this explains why both their train and test MSE are higher than the less complex models. Regardless, it seems that for up to about 200 total nodes, three hidden layers does the best, and as such we chose an architecture of three hidden layers with 66 nodes each (which is where we achieved our lowest MSE).

\begin{figure}[h]
\centerline{\includegraphics[width=6in]{figure/msebymodelfranke.png}}
\caption{Comparison between a neural net with one hidden layer, and one with two hidden layers, for different totals of hidden nodes. For the network with two hidden layers, the total is evenly divided between the layers, such that at $x = 60$ we have a network with two hidden layers of 30 nodes each. In this case, the hidden layers used the sigmoid activation function.}
\label{msebymodelfranke}
\end{figure}

Using this architecture, we performed a gridsearch. Due to time constraints, we have felt it necessary to limit our gridsearch to only two schedulers. Based on our testing with linear regression with gradient descent, we have seen that Adam and Adagrad seems to be the best performers. Based on Goodfellows recommendations, we started with the RELU activation function \cite[188]{gbc}. With this, we found Adam to be the best scheduler, with an eta of 0.01 and no regularization. However, when using these parameters, we encountered the so called "dying RELU problem", where too many of our RELU nodes output zero, leading to our gradient descent slowing down and eventually stopping completely. A better weight initialization, for example using Xavier Initilization, could solve this problem, but another way to circumvent this issue is to use Leaky RELU. This activation function has a small but non-zero gradient for negative values, instead of zero as with RELU. Gridsearching with Leaky RELU instead gave the exact same parameters.

Using these parameters along with Adam and LRELU, we acheived and MSE of VERYNICE, which is better than the results gained using gradient descent on our polynomial fit. Furthermore, we beat our gradient descended polynomial fit after a very short amount of epochs, achieving MSEs lower than 0.011 (the best gradient descended polynomial fit) after only EPOCH epochs. We can see that by not forcing our model to try and find a polynomial fit, we are achieving better results more quickly. However, we are still not able to beat the analytical solution to the best polynomial fit at polynomial degree 9, which is somewhat disappointing. A possible explanation for this is that the Franke function can be represented as a weighted sum of polynomials, and as such a polynomial fit is especially suited to approximate this function. Our neural net must learn the relation between the inputs and outputs on its own, a process which is computationally intensive and difficult. Looking at our predicted terrain (Fig.~\ref{predterrain}), we can see that our prediction is somewhat reasonable, but not entirely convincing.

\begin{figure}[h]
\centerline{\includegraphics[width=6in]{figure/Franke_fit_66x3_lam0.png}}
\caption{Comparison of our prediction of }
\label{predterrain}
\end{figure}

@@

With mediocre results on regression, we turn to classification. To test our neural networks ability to solve classification problems, we first applied it to the Wisconsin Breast Cancer dataset. But to get the best results, there are a large amount of parameters that need to be decided on. We must find the best architecture, the best activation functions, best scheduler and the best parameters for the scheduler. Finding the absolute best combination of all these choices is impossible, so compromises must be made.

We start by trying to find a good model complexity, i.e the best number of hidden layers, and the amount of nodes in them. To do this, we plotted the accuracy over different model architectures. Like we did during regression, we fix all the other parameters and study the architecture by itself. Using the sigmoid activation function (Fig.~\ref{cancerarchcompsigmoid}), our results are somewhat surprising. We see that even at 400 total hidden nodes, we are not overfitting. This implies that the dataset is quite homogenous, and that the test set is very similar to our training set. With such a dataset, it is possible to gain a very good fit using more and more complex models, although the computational effort obviously increases as well. Moreover, it is not given that instances outside this dataset are as homogenous as the relatively small (569) sample we have here, so one should be careful to use an overcomplex model even if it does not hurt the test accuracy for this particular dataset. Using the RELU activation function, the results are quite different, as can be seen in Fig.~\ref{cancerarchcomp}. Here we can observe that a neural net with two or three hidden layers starts performing very inconsistently when the number of nodes increases. Looking more closely at the training process, we can see that training stops completely in some of our crossvalidations. Clearly, we are again running into the "dying RELU problem". Here we can clearly see some of the benefits of the sigmoid, as even if the gradients become very small, they will never be zero, as with RELU. Regardless, we see similar performance for RELU using only one hidden layer. Based on these findings, we decided to use a model with a single hidden layer of 100 nodes, using the RELU activation function.

\begin{figure}[h]
\centerline{\includegraphics[width=6in]{figure/cancerarchcompsigmoid.png}}
\caption{Comparison between a neural net with one hidden layer, and one with two hidden layers, for different totals of hidden nodes. For the network with two hidden layers, the total is evenly divided between the layers, such that at $x = 60$ we have a network with two hidden layers of 30 nodes each. In this case, the hidden layers used the sigmoid activation function.}
\label{cancerarchcompsigmoid}
\end{figure}

\begin{figure}[h]
\centerline{\includegraphics[width=6in]{figure/cancerarchcomp.png}}
\caption{Comparison between a neural net with one hidden layer, and one with two hidden layers, for different totals of hidden nodes. For the network with two hidden layers, the total is evenly divided between the layers, such that at $x = 60$ we have a network with two hidden layers of 30 nodes each. In this case, the hidden layers used the RELU activation function.}
\label{cancerarchcomp}
\end{figure}

Using our gridsearch method on this model, we found that the best scheduler was Adam, with an eta of $0.001$ and a lambda of $0.0001$. With this combination of architecture and scheduler, we were able to achieve a crossvalidated test accuracy of $98.76\%$, which is a great result. This outperforms SciKit's MLP classifier, which achieved a crossvalidated test accuracy of around 95\%. Looking at the confusion matrix (Fig.~\ref{cancerconf}), we see that we are achieving good predictions for both true positives and true negatives.

\begin{figure}[h]
\centerline{\includegraphics[width=5in]{figure/cancerconf.png}}
\caption{Confusion matrix for the cancer dataset using a neural net with a single hidden layer of 60 nodes}
\label{cancerconf}
\end{figure}

Confident in our networks's ability to do binary classification, we turned to the MNIST dataset. Here, we perform an educated guess on the different parameters, based on our previous testing. We decided upon an architecture of a single layer with 100 nodes using the RELU activation function. For our scheduler, we used Adam with an eta of $0.001$ and no regularization. Being a multiclass problem with ten different classes, one might expect our neural network to struggle more with this dataset. However, this was not the case, as we achieved a crossvalidated test accuracy of $99.07\%$. Based on these results, we conclude that the initial guess for our model was a good one, and that further optimization has little room for improvement. From these two datasets, we can see that a neural network is adept at solving classification problems. On the other hand, these good results, and the ease with which our network attained them, could imply that the problems we have dealt with are not as complicated as they may seem.

@@

In this paper we have explored the use of a Feed Forward Neural Network for both regression and classification. We have studied how different schedulers, affect the gradient descent

@@

\begin{table*}
\caption{This table shows the file which created every plot in the report}
\begin{center}
\label{allparamstable}
\begin{tabular}{c | l l l}
Figure & Shell command \\
\hline
\ref{verification} & \texttt{python3 testing/verification.py}\\
\ref{cancerconflog} & \texttt{python3 task\_d.py}\\
\ref{cancerconf} & \texttt{python3 task\_d.py}\\
\ref{cancerlogneur} & \texttt{python3 task\_d.py}\\
\ref{cancerarchcomp} & \texttt{python3 bycomplexitycancer.py}\\
\ref{cancerarchcompsigmoid} & \texttt{python3 bycomplexitycancer.py}\\
\ref{msebymodelfranke} & \texttt{python3 bycomplexityfranke.py}\\
\hline
\end{tabular}
\end{center}
\end{table*}
